---
title: Hallmarks GSEA
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center', 
                      fig.width = 9, fig.height = 6.5)
```

<!--
//TODO  -When having two scores with the same name its average was used. 
        -Zeros were removed. 
        -Scores without names (which can not be in any signature) removed. 
        -Non complete cases (i.e. NAs, NaNs) were removed
        -Doesn't work when only given a single geneset
-->


```{r}
library(ggplot2)
library(tidyverse)
library(ggplot2)
library(GSEABase)
library(phenoTest)
library(testthat)
library(ggthemes)
```

```{r}
edset <- read_rds("../output/edset-organised.rds")
```

#### Here we're going to compare enrichment of the [Hallmark genesets](https://www.ncbi.nlm.nih.gov/pubmed/26771021) between dormant and desensitised patients.

### Functions

#### Geneset arrangement

```{r}
arrangeGenesets <- function(GMT){
    genesets <- geneIds(GMT)
    intersets <- lapply(genesets, function(set){
                            intersect(set, row.names(edset))
    })
    intersets
}
```

#### GSEA calculation functions (for unpaired and paired comparisons)

```{r}
arrangeGSEA <- function(gsea_out, comparison = NA, dormancy = NA){
    dfr <- data.frame(summary(gsea_out)[, c("nes", "fdr")])
    dfr[dfr$fdr >= 0.05, "nes"] <- NA
    dfr$hallmark <- row.names(dfr)
    dfr$comparison <- comparison
    dfr$dormancy <- dormancy
    dfr$direction <- ifelse(grepl("UP$", dfr$hallmark), "UP", 
           ifelse(grepl("DN$", dfr$hallmark), "DOWN", "EITHER"))
    dfr
}
```

#### Unpaired (i.e. between class) GSEA calculation

```{r}
arrangeUnpairedGSEA <- function(eset, geneset){
    #separate into dormants and desensitiseds
    dorm_xpr <- exprs(eset[, which(eset$is_dormant)]) %>% rowMeans()
    dssd_xpr <- exprs(eset[, which(!eset$is_dormant)]) %>% rowMeans()
    #calculate fold-change - this orientation is relative to desensitiseds
    fc_vec <- sort(dssd_xpr - dorm_xpr)

    test_that("test that our vector meets the phenoTest reccomended specs", {
        expect_false(unique(fc_vec %>% is.na()))
        expect_false(unique(fc_vec == 0))
        #even at 15 decimals points, for all genes there are value ties but they're such a small
        #percentage of all genes that I don't think it is critical to the results here
    })
    #perform GSEA and arrange for plotting
    geneset_out <- gsea(x = fc_vec, gsets = geneset, logScale = FALSE)
    arrangeGSEA(geneset_out, comparison = "Dormant-Desensitised")
} 
```

#### Plot the unpaired calculations, both as a whole but also split by treatment timepoint

```{r}
plotUnpaired <- function(x){
    ggplot(x, aes(x = comparison, y = hallmark, fill = nes)) + 
            geom_tile(colour = "white") +
            scale_fill_gradientn(values=c(1, .7, .5, .3, 0), colours=rev(c("#4eb3d3", "#a8ddb5", "white", "#fee391", "#fe9929")), na.value = "white") +
            labs(title = "GSEA : Dormant vs. Desensitised", subtitle = "Enrichment is relative to acquired resistant patients") 
}

plotUnpairedSplit <- function(geneset){ 
        plot_dfr <- lapply(c("diagnostic", "early", "extended"), function(x) {
        dfr <- arrangeUnpairedGSEA(edset[, which(edset$treatment == x)], geneset) 
        dfr$timepoint <- factor(x, levels = c("diagnostic", "early", "extended"))
        dfr
        }) %>% do.call(rbind, .) 
        ggplot(plot_dfr, aes(x = timepoint, y = hallmark, fill = nes)) +
            geom_tile(colour = "#f0f0f0") + 
            scale_fill_gradientn(values=c(1, .7, .5, .3, 0), colours=rev(c("#4eb3d3", "#a8ddb5", "white", "#fee391", "#fe9929")), na.value = "white") +
            scale_x_discrete(labels = c("Diagnostic", "Early-on", "Extended")) + 
            theme_pander()
} 
```


### Analysis

#### First we'll assemble the Hallmark and cancer-specific genesets, as a sensible easy to use list of ids that are present in `edset`.

```{r}
hallmarks <- getGmt("../../combined/data/h.all.v6.1.symbols.gmt") %>% arrangeGenesets()
names(hallmarks) <- gsub("HALLMARK ", "", gsub("_", " ", names(hallmarks)))
```

#### We'll begin with an unpaired comparison, by arranging our data as two sets - dormants & desensitiseds - calculating the significant **NES** scores, and plotting.

### Plots

```{r, fig.width = 5.5}
#hallmarks
hallmarks_out <- arrangeUnpairedGSEA(edset, hallmarks)
hallmarks_out %>%
        plotUnpaired()
```

#### We can then split our dormant vs. desensitised comparison by timepoint

```{r}
plotUnpairedSplit(hallmarks)
```

