---
title: Differential expression
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')
```

```{r}
library(tidyverse)
library(Biobase)
library(siggenes)
library(RankProd)
library(reshape2)
library(ggthemes)
```

```{r}
dorm <- read_rds("../output/dorm-v4.rds")
```

# Within groups

## Dormants
- remove non-dormants and timepoint 1 samples
- find sample pairs
- if there are two timepoint 2 samples only take the early (e) version 
 
```{r}
dorm_d <- dorm[, dorm$dorm.group_v3 == "D" & dorm$time.point_3cat != 1]
pairs_d24 <- which(colnames(dorm_d) %>% gsub("\\-[1-9]", "", .) %>% table() == 2) %>% names() %>% paste0("^", .)
dorm_d24_dbls <- dorm_d[, sapply(pairs_d24, function(x) grep(x, colnames(dorm_d))) %>% unlist()]
dorm_d24 <- dorm_d24_dbls[, -c(grep("2e", colnames(dorm_d24_dbls)) - 1, grep("4e", colnames(dorm_d24_dbls)))]
```

### Calculate differential expression

```{r}
cl_d24 <- rep(c(-1, 1), ncol(dorm_d24)/2) * rep(1:(ncol(dorm_d24)/2), each = 2)

sam_d24 <- sam(exprs(dorm_d24), cl_d24, rand = 221)
findDelta(sam_d24, 0.05) # 1.132684
geneset_d24 <- list.siggenes(sam_d24, 1.132684)

dorm_d24sig <- dorm_d24[geneset_d24,]
upreg <- (exprs(dorm_d24sig)[, seq(2, 58, 2)] - exprs(dorm_d24sig)[, seq(1, 58, 2)]) %>% rowMeans() > 0
upreg_dfr <- data.frame(Var1 = names(upreg), is.up = upreg)
```

### I think this was a not-so-smart way to do it = IGNORE

```{r, eval = FALSE, echo = FALSE}
dorm_mlt <- melt(exprs(dorm[geneset_d24,]))
dorm_mrg_base <- merge(dorm_mlt, upreg_dfr, by = "Var1")
dorm_mrg <- merge(dorm_mrg_base, pData(dorm), by.x = "Var2", by.y = "ID_D120days")

#the dimensions of mlt and mrg potential don't match but too fuzz brained to figure it out CHECK!!!!!!!

ggplot(dorm_mrg, aes(x = as.character(time.point_3cat), y = value)) + 
    geom_line(aes(group = patient.no, colour = is.up)) + 
    geom_boxplot(alpha = 0.7) + 
    facet_grid(is.up~dorm.group_v3) + 
    theme_pander()

ggplot(dorm_mrg, aes(x = Var2, y = Var1, fill = value)) +
    geom_tile() +
    facet_wrap(~dorm.group_v3, scales = 'free_x', ncol = 1)
```

### Calculate for row-wise and column-wise mean expression and then plot, facetting by dormancy group and differential expression direction

#### Row-wise (i.e. per-gene means)

```{r}
genemean_dfr <- lapply(c("D", "ND"), function(class){
    subdorm <- dorm[, dorm$dorm.group_v3 == class]
    lapply(c("\\-1", "\\-2", "\\-4"), function(x) {
                           row_means <- exprs(subdorm[geneset_d24, grep(x, colnames(subdorm))]) %>% rowMeans()
                           data.frame(gene = names(row_means), xpr =row_means, timepoint = substring(x, 3, 3), class = class)
    }) %>% do.call(rbind, .)
}) %>% do.call(rbind, .)
genemean_mrg <- merge(genemean_dfr, upreg_dfr, by.x = "gene", by.y = "Var1")
genemean_mrg$direction <- ifelse(genemean_mrg$is.up == TRUE, "Up", "Down")

ggplot(genemean_mrg, aes(x = timepoint, y = xpr)) + 
    geom_line(aes(group = gene, colour = class), alpha = 0.2) + 
    geom_boxplot(alpha = 0.7, notch = TRUE) + 
    facet_grid(direction~class) + 
    scale_colour_manual(values = c("#ff6f69", "#5bc0de")) +
    labs(title = "Dormant DE genes 2-4 - genewise", 
         subtitle = "Mean (per-gene) expression of DE genes between On- and Long-term timepoints were\ncalculated in Dormant patients and then applied to both dormancy classes\nacross all timepoints",
         x = "Biopsy",
         y = "Mean Expression") + 
    theme_pander() + 
    theme(legend.position = 'none',
          panel.background = element_blank(), 
          panel.grid.major = element_line(colour = "#d9d9d9", linetype = 1),
          panel.grid.minor = element_line(colour = "#d9d9d9", linetype = 1))
```

#### Column-wise (i.e. per-sample means)

```{r}
it_lst <- list(Up = row.names(upreg_dfr[which(upreg_dfr$is.up),]), Down = row.names(upreg_dfr[-which(upreg_dfr$is.up),]))
sampmean_dfr <- lapply(c("Up", "Down"), function(direction){
    col_means <- exprs(dorm[it_lst[[direction]],]) %>% colMeans()
    data.frame(id = names(col_means), xpr = col_means, direction  = direction)
}) %>% do.call(rbind, .)
sampmean_mrg <- merge(sampmean_dfr, pData(dorm), by.x = "id", by.y = "ID_D120days_3cat")

ggplot(sampmean_mrg, aes(x = time.point_3cat, y = xpr)) + 
    geom_line(aes(group = patient.no, colour = dorm.group_v3)) + 
    geom_boxplot(alpha = 0.7, notch = TRUE) + 
    facet_grid(direction~dorm.group_v3) + 
    scale_colour_manual(values = c("#ff6f69", "#5bc0de")) +
    scale_x_discrete(labels = c("Pre", "On", "Long-term")) + 
    labs(title = "Dormant DE genes 2-4 - samplewise", subtitle = "Mean (per-sample) expression of DE genes between On- and Long-term\ntimpoints were calculated in Dormant patients and then applied to both\ndormancy classes across all timepoints",
    x = "Biopsy",
    y = "Mean Expression") + 
    theme_pander() + 
    theme(legend.position = 'none',
          panel.background = element_blank(), 
          panel.grid.major = element_line(colour = "#d9d9d9", linetype = 1),
          panel.grid.minor = element_line(colour = "#d9d9d9", linetype = 1))
#ggsave("../../../writing/PhD/3yr/figs/pres/de-d24.png")
```

## Desensitiseds

### Organise as above
 
```{r}
dorm_n <- dorm[, dorm$dorm.group_v3 == "ND" & dorm$time.point_3cat != 1]
pairs_n24 <- which(colnames(dorm_n) %>% gsub("\\-[1-9]", "", .) %>% table() == 2) %>% names() %>% paste0("^", .)
dorm_n24_dbls <- dorm_n[, sapply(pairs_n24, function(x) grep(x, colnames(dorm_n))) %>% unlist()]
dorm_n24 <- dorm_n24_dbls[, -c(grep("2e", colnames(dorm_n24_dbls)) - 1, grep("4e", colnames(dorm_n24_dbls)))]
```

### Calculate differential expression

```{r}
cl_n24 <- rep(c(-1, 1), ncol(dorm_n24)/2) * rep(1:(ncol(dorm_n24)/2), each = 2)

sam_n24 <- sam(exprs(dorm_n24), cl_n24, rand = 221)
```

# Across groups

## SAM


```{r}
dorm_4 <- dorm[, dorm$time.point_3cat == 4]

cl <- ifelse(dorm_4$dorm.group_v3 == "D", 0, 1)

sam_out <- sam(exprs(dorm_4), cl, rand = 221)
```

#### No genes differentially expressed at FDR<=0.05 by SAM


## RankProduct

```{r, eval = FALSE}
RP_out <- RP(exprs(dorm_4), cl, logged=TRUE, na.rm=FALSE, plot=FALSE, rand=123)
```

```{r}
topgenes <- topGene(RP_out, cutoff = 0.05)

geneset <- row.names(dorm_4)[data.frame(do.call(rbind, topgenes))$gene.index]

#write_lines(geneset, "../output/rp-geneset.txt")
```

