---
title: Batch effects and correction
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center', fig.width = 8, fig.height = 6)
```

#### *NOTE: these normalisation and batch correction files have been run recursively, i.e. once with all samples to determine which duplicates to remove, then with duplicates removed, then with `is_dormant == NA` samples removed. They will therefore only detail the final run-through though all the code from raw to final expressionsets remains, only the input files need to be changed.*

```{r}
library(tidyverse)
library(Biobase)
library(sva)
library(reshape2)
library(testthat)
library(ggthemes)
library(cowplot); theme_set(theme_gray())
source("../../../../functions/mostVar.R")
source("../../../../functions/library/mdsArrange.R")
```

#### We have our frma and loess normalised expression set, but there still clearly exists a big batch difference between FF and FFPE source material. This is evident in terms of the overall sample distributions

```{r}
georgeset <- read_rds("../output/georgeset-sep-frma-fselect-loess-clin.Rds")
```

## Visual determination

### Pre-correction

```{r}
batchEffectDists <- function(mtx_input){
source("../../../../functions/mostVar.R")
source("../../../../functions/library/mdsArrange.R")
    #arrange
    george_mlt <- melt(mtx_input)
    george_mrg <- base::merge(george_mlt, pData(georgeset), by.x = "Var2", by.y = 0)
    george_mrg$Var2 <- factor(george_mrg$Var2, levels = unique(george_mrg$Var2[order(george_mrg$is_freshfrozen)]))
    #boxplot
    p_box <- ggplot(george_mrg, aes(x = Var2, y = value, fill = is_freshfrozen)) + 
        geom_boxplot() +
        theme(legend.position = 'bottom',
            axis.ticks.x = element_blank(),
            axis.text.x = element_blank())
    #densityplot
    p_dist <- ggplot(george_mrg, aes(x = value, colour = is_freshfrozen, group = Var2), alpha = 0.05) + 
        geom_line(stat = 'density', alpha = 0.2) + 
        theme_pander() +    
        theme(legend.position = 'bottom',
            axis.ticks.x = element_blank(),
            axis.text.x = element_blank()) 
        #combine
    plot_grid(p_box, p_dist, ncol = 1)
}

batchEffectDists(exprs(georgeset))
```

#### ...as well as when comparing just the 500 most variable genes as an mds

```{r}
batchEffectMDS <- function(mtx_input, colour_by){
    #arrange
    mv500 <- mostVar(mtx_input, 500) %>% row.names()
    arg500 <- mdsArrange(mtx_input[mv500,]) 
    mds_input <- base::merge(arg500, pData(georgeset), by.x = 'ids', by.y = 0)
    #and plot
    ggplot(mds_input, aes_string("x", "y", colour = colour_by)) + 
        geom_point() + 
        theme_pander() + 
        theme(legend.position = 'bottom')
}

batchEffectMDS(exprs(georgeset), "is_freshfrozen")
```

#### We can additionally check a number of other variables to check for any other effects

```{r}
var_vec <- c("biopsy", "is_uss", "Age", "ER", "Diagnostic.CB.Grade", "No.of.Pos.Nodes")
plot_lst <- lapply(var_vec, function(variable) batchEffectMDS(exprs(georgeset), variable))
cowplot::plot_grid(plotlist = plot_lst, ncol = 3)
```

#### And as it's perhaps the most likely source of biological variation, we'll plot biopsy number a little larger

```{r}
batchEffectMDS(exprs(georgeset), "timepoint")
```

### Correction and post-correction

#### We'll attempt to correct for this source material-specific batch effect using ComBat

```{r}
g_cb <- ComBat(exprs(georgeset), batch = as.numeric(georgeset$is_freshfrozen))
```

#### And XPN

*note, I've ignored XPN here as it was a bit hacky to get it working and I don't fully trust it*

```{r, eval = FALSE}
#lapply(list.files("~/Desktop/temp-store/inSilicoMerging/R/"), function(file){
#           print(file)
#           source(file.path("~/Desktop/temp-store/inSilicoMerging/R", file))
#            })
#source("~/Desktop/temp-store/inSilicoMerging/R/xpn.R")
#source("~/Desktop/temp-store/inSilicoMerging/R/util.R")
#
##g_xpn <- merge(list(georgeset[, georgeset$is_freshfrozen], georgeset[, !georgeset$is_freshfrozen]), method = "XPN")
#aux <- xpn(georgeset[, georgeset$is_freshfrozen], georgeset[, !georgeset$is_freshfrozen])
#g_xpn <- a4Base::combineTwoExpressionSet(aux[[1]], aux[[2]])
##write_rds(aux, "~/Desktop/temp-store/gset-frma-feature-loess-xpn.rds")
```

ComBat

```{r}
batchEffectDists(g_cb)
```

#### Better? What about by MDS?

ComBat

```{r}
batchEffectMDS(g_cb, "is_freshfrozen")
batchEffectMDS(g_cb, "timepoint") 
```

#### Prior to feature selecting, the FF vs FFPE effect was still very evident post-correction but now the clear divide caused by source material has seemingly all but been removed. Therefore, we'll write out the final expression set to be used in the actual analysis

```{r}
georgeset_cb <- georgeset
exprs(georgeset_cb) <- g_cb

#test we haven't broken our expressionset as can happen!
test_that("eset overwrite worked as expected", {
              expect_identical(g_cb, exprs(georgeset_cb))
             expect_identical(pData(georgeset), pData(georgeset_cb))
           })

write_rds(georgeset_cb, "../output/final-georgeset-sep-frma-fselect-loess-clin-cb.Rds")
```

### *The following is now redundent as duplilcate samples have been determined and removed previously*

#### Let's now check the FF-FFPE replicate samples only 

```{r, eval = FALSE}
batchEffectDists(g_cb[, georgeset$is_material_rep])
batchEffectMDS(g_cb[, georgeset$is_material_rep], "sample_id") + geom_point(aes(size = is_freshfrozen))
```


## Statistical determination 

We can losely define the batch-sepcific *inter*-patienty and *intra*-patient heterogeneity as biological variation, just as the replicate sample variation can be almost entirely attributed to technical variation or noise. Additionally, the overall variability will be explained by a combination of the above.

Therefore, post-ComBat, if the 'biological' variation remains constant post-ComBat, whislt the technical variation is removed we can be confident that our correction has been effective but not to the degree of removing bioloical signal.

```{r, eval = FALSE}
getSampleCors <- function(mtx){
    cor_mtx <- cor(mtx)
    diag(cor_mtx) <- NA
    colMeans(cor_mtx, na.rm = TRUE)
}

input_lst <- list(raw_all = exprs(georgeset), 
                  raw_ff = exprs(georgeset)[, georgeset$is_freshfrozen],
                  raw_ffpe = exprs(georgeset)[, !georgeset$is_freshfrozen],
                  raw_rep = exprs(georgeset)[, georgeset$is_material_rep],
                  cb_all = exprs(georgeset_cb), 
                  cb_ff = exprs(georgeset_cb)[, georgeset_cb$is_freshfrozen],
                  cb_ffpe = exprs(georgeset_cb)[, !georgeset_cb$is_freshfrozen],
                  cb_rep = exprs(georgeset_cb)[, georgeset_cb$is_material_rep]
                  )

cor_dfr <- lapply(names(input_lst), function(x){
                      data.frame(cor = getSampleCors(input_lst[[x]]), 
                                 comparison = x, 
                                 xpr_id = colnames(input_lst[[x]]))
                  }) %>% do.call(rbind, .)
cor_dfr$correction <- factor(ifelse(grepl("raw", cor_dfr$comparison), "RAW", "CB"), 
                             levels = c("RAW", "CB"))

ggplot(cor_dfr, aes(x = comparison, cor)) + 
    geom_boxplot(outlier.size = 0) + 
    geom_jitter(width = 0.1, alpha = 0.7) +
    facet_wrap(~correction, nrow = 1, scales = 'free_x') + 
    labs(x = "Comparison Type", 
         y = "Correlation", 
         title = "Pre- and post-correction correlation") +
    theme_pander() +
    theme(panel.spacing = unit(5, "lines"))
```

Happily we can see that whilst our overall and replicate correlations have drastically improved, this has not been at the expense of our batch-specific correlations - i.e. biological variation has been preserved.

```{r, eval = FALSE}
lapply(names(input_lst), function(x){
           data.frame(Correlation = mean(cor_dfr[cor_dfr$comparison == x,]$cor), 
                      Comparison = x) 
         }) %>% 
    do.call(rbind, .) %>%
    t() %>% 
    knitr::kable()
```

## Duplicate samples selection and removal

Finally, we can use our correlations to determine which samples to remove and which to keep. Our decision here will be made as two parts: 1. if the sample is FF, keep else 2. keep the FFPE replicate with the highest pre-correction intra-batch correlation

```{r, eval = FALSE}
ffpecor_dfr <- cor_dfr[which(cor_dfr$comparison %in% c("raw_ffpe")),]
decision_dfr <- merge(ffpecor_dfr, pData(georgeset), by = "xpr_id")

materialreps_toremove <- decision_dfr[decision_dfr$is_material_rep & !decision_dfr$is_freshfrozen,
                                      "xpr_id"] %>% as.character()

ffpedecision_dfr <- decision_dfr[decision_dfr$is_replicate, ]
ffpe_rep_patients <- unique(ffpedecision_dfr$sample_id)


technicalreps_toremove <- lapply(ffpe_rep_patients, function(sample){
           dfr <- ffpedecision_dfr[which(ffpedecision_dfr$sample_id == sample),]
           dfr$xpr_id[-which.max(dfr$cor)] %>% as.character
         }) %>% unlist()
```

With these replicates to remove we physically do this in the intitial .CEL file directory, before AffyBatch creation

```{r, eval = FALSE}
c(technicalreps_toremove, materialreps_toremove) %>%
    unique() %>% 
    sort() %>% 
    cat(sep = "\n")
```


