---
title:
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')
```

```{r}
library(tidyverse)
library(Biobase)
library(affy)
library(frma)
library(hgu133plus2frmavecs)
library(a4Base)
```

The clinical information is a mess so we need to do some extensive curation and tidying. We'll
start with the USS data and begin by filling in missing values.

## USS

```{r}
uss_dfr <- read.csv("../data/batch-2-uss.csv")
```

For `Patient.ID` we need to simply replace NAs with the first previous value. So we take the
indexes where the values are, calculate the space inbetween them and fill that space with the
correct value. Note we add an additional value which is the number of rows + 1 to deal with the
final edge case - we therefore only loop through the length of idx - 1.

We'll wrap this in a function as we'll use the same approach for the `Sample.ID` column.

```{r}
fillNAs <- function(vec){
    vec = as.character(vec)
    idx <- c(which(!is.na(vec)), nrow(uss_dfr) + 1)
    fill_vec <- rep(NA, nrow(uss_dfr))
    for(i in 1:(length(idx)-1)){
        start <- idx[i]
        stop <- diff(c(idx[i], idx[i+1])) + idx[i] - 1
        fill_vec[start:stop] <- vec[start]
    }
    fill_vec
}

uss_dfr$patient_id <- fillNAs(uss_dfr$Patient.ID)
#uss_dfr$sample_id <- fillNAs(uss_dfr$Sample.ID) #actually this doesn't make sense to do
```

Now we add two columns, specifiying whether the row corresponds to a biopsy, a USS measurement or 
both. In addition we add another column detailing which time point each biopsy corresponds to.

```{r}
uss_dfr$is_uss <- ifelse(is.na(uss_dfr$Relative.Volume), FALSE, TRUE)
uss_dfr$is_biopsy <- ifelse(is.na(uss_dfr$Surgical.Excision..TH..or.Core.Biospy..CB.), FALSE, TRUE)

uss_dfr$timepoint <- ifelse(uss_dfr$Days.from.Treatment <= 0,
                            "diagnosis",
                            ifelse(uss_dfr$Days.from.Treatment <= 120,
                                   "on-treatment", 
                                   "long-term"))
```

Finally we delete now obsolete columns, rename any which are needed but haven't yet been modified 
and re-order columns

```{r}
uss_dfr[, c("Patient.ID", "Surgical.Excision..TH..or.Core.Biospy..CB.")] <- NULL

uss_dfr <- uss_dfr[, c("patient_id", "Sample.ID", "timepoint", "is_biopsy", 
                       "is_uss", "Relative.Volume", "Days.from.Treatment", "Date")]
colnames(uss_dfr) <-  c("patient_id", "sample_id", "timepoint", "is_biopsy", 
                        "is_uss", "relative_vol", "days_treated", "date")
```

# Clinical

```{r}
clin_dfr <- read.csv("../data/batch-2-clin.csv")
```

Not too many changes to make here, only patient 283 in `clin_dfr` is called 283R in `uss_dfr` so 
we'll remove the Rs before merging both in `patient_id` and `sample_id` columns.

```{r}
uss_dfr$patient_id[grepl("R", uss_dfr$patient_id)] <- "283"
uss_dfr$sample_id <- lapply(as.character(uss_dfr$sample_id), function(x){
                                ifelse(grepl("R", x), gsub("R", "", x), x)
                        }) %>% unlist()

ussclin_dfr <- merge(uss_dfr, clin_dfr, by.x = 'patient_id', by.y = 'LA.Number', all = TRUE)
```

# Expression

Lastly we can add to our pheno data by parsing information from the sample names, before combining
expression and pheno data as a complete expression set

```{r}
##read .CEL files
#setwd("../data/Edinburgh\ Second\ Batch\ 170\ Samples")
#batch_2 <- ReadAffy()
#setwd("../../src/")

##feature selection
ap <- mas5calls(batch_2)

ap_ffpe <- exprs(ap)[, which(!grepl("FF", colnames(ap), ignore.case = TRUE))]
ap_ff <- exprs(ap)[, which(grepl("FF", colnames(ap), ignore.case = TRUE))]

calls_ffpe <- sapply(row.names(ap), function(x){
    sum(ap_ffpe[x, ] != "A") > 10
})
present_ffpe <- row.names(ap)[which(calls_ffpe)]
 
calls_ff <- sapply(row.names(ap), function(x){
    sum(ap_ff[x, ] != "A") >= 7
})
present_ff <- row.names(ap)[which(calls_ff)]

present_common <- intersect(present_ffpe, present_ff)

data.frame(present_10 = c(length(present_ffpe), 
                             length(present_ff), 
                             length(present_common)),
            material = c("FFPE", "FF", "Common")
            ) %>% knitr::kable()

write_lines(present_common, "../output/probes-present-common.txt")

##2-step normalisation - frma and loess 
#data(hgu133plus2frmavecs)
#georgeset <- frma(batch_2, input.vecs = hgu133plus2frmavecs) 

#g_mtx <- affy::normalize.loess(exprs(georgeset))
#write.csv(g_mtx, "../output/loess-norm-xpr.txt")

##this was a weird error about mismatched dimnames but fixed by setting colnames explicitly
#colnames(georgeset) <- colnames(exprs(georgeset))
#exprs(georgeset) <- g_mtx
#write_rds(georgeset, "../output/batch-2-frma-loess.Rds")
georgeset <- read_rds("../output/batch-2-frma-loess.Rds")
```

This requires some complicated and careful calls : double-check, double-check, double-check

```{r}
georgeset$is_freshfrozen <- ifelse(grepl("FF", colnames(georgeset), ignore.case = TRUE),
                                   TRUE,
                                   FALSE)

georgeset$is.control <- ifelse(grepl("VHRR", colnames(georgeset), ignore.case = TRUE), 
                                   TRUE, 
                                   FALSE)

#the largest sample id is in the hundreds so chop to 3 characters and then remove anything that's 
#not a number
tmp_id1 <- substring(colnames(georgeset), 1, 3) 
georgeset$patient <- gsub("\\W", "", tmp_id1)

#remove everything to the first `-` and then take the next character
tmp1 <- gsub("\\d+\\w-", "", colnames(georgeset), ignore.case = TRUE)
tmp2 <- substring(tmp1, 1, 1)
georgeset$biopsy <- tmp2
georgeset$biopsy[georgeset$biopsy == "V"] <- NA

georgeset$sample_id <- paste(georgeset$patient, georgeset$biopsy, sep = '-')
georgeset$sample <- NULL

georgeset$xpr_id <- colnames(georgeset)

georgeset$platform <- "HG-U133_Plus_2"

#this needs work
georgeset$is.replicate <- ifelse(grepl("-([0-9])-([0-9])([a-z])_", colnames(georgeset),
                                       ignore.case = TRUE), 
                                   TRUE, 
                                   FALSE)

pData(georgeset) <- lapply(unique(georgeset$patient), function(x){
           subset_dfr <- pData(georgeset)[which(georgeset$patient == x),]
           ff_tbl <- subset_dfr$is_freshfrozen %>% table()    
           if(length(ff_tbl) == 2){
               subset_dfr$is_material_rep = TRUE
           } else {
               subset_dfr$is_material_rep = FALSE
           }
           subset_dfr
}) %>% do.call(rbind, .) 
```

Merge the final version of the phenotypic data and create the final expressionSet

```{r}
pheno <- merge(ussclin_dfr, pData(georgeset), by = 'sample_id', all.y = TRUE)
pheno$has_clin <- ifelse(pheno$is_biopsy, TRUE, FALSE)

pData(georgeset) <- pheno
#write_rds(georgeset, "../output/georgeset.rds")
```


