---
title: Georgetown Pre-processing
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')
```

```{r}
library(affy)
library(a4Base)
library(tidyverse)
library(knitr)
library(ggthemes)
source("../../../../functions/mostVar.R")
source("../../../../functions/library/mdsArrange.R")
```

## First we'll read in our .cel files in batch, normalise using RMA, add batch information, combine, edit the samples names to remove uneeded information and finally add fresh-frozen infromation.

## Actually one more thing, there appear to be technical replicates and control samples so we'll highlight this as well

```{r, eval = FALSE}
setwd("../data/Edinbrugh\ First\ Batch\ 170\ Samples")
batch_1 <- ReadAffy()
setwd("../Edinburgh\ Second\ Batch\ 170\ Samples/")
batch_2 <- ReadAffy()
setwd("../../src/")
batch_lst <- list(batch_1, batch_2)

eset_lst <- lapply(1:2, function(x){
    eset <- rma(batch_lst[[x]])
    eset$batch <- LETTERS[x]
    eset
})

#georgeset <- do.call(combineTwoExpressionSet, eset_lst)
georgeset <- eset_lst[[2]]

georgeset$is.freshfrozen <- ifelse(grepl("FF", colnames(georgeset), ignore.case = TRUE),
                                   TRUE, 
                                   ifelse(georgeset$batch == "B", 
                                          FALSE, 
                                          NA))

georgeset$is.replicate <- ifelse(grepl("-([0-9])-([0-9])([a-z])_", colnames(georgeset),
                                       ignore.case = TRUE), 
                                   TRUE, 
                                   FALSE)

georgeset$is.control <- ifelse(grepl("VHRR", colnames(georgeset), ignore.case = TRUE), 
                                   TRUE, 
                                   FALSE)

#the largest sample id is in the hundreds so chop to 3 characters and then remove anything that's 
#not a number
tmp_id1 <- substring(colnames(georgeset), 1, 3) 
georgeset$patient <- gsub("\\W", "", tmp_id1)

tmp1 <- gsub("[0-9]+-", "", colnames(georgeset))
tmp2 <- substring(tmp1[georgeset$batch == "B"], 1, 1)
tmp3 <- ifelse(grepl("PRE", colnames(georgeset)[georgeset$batch == "A"]), 1, 4)
georgeset$biopsy <- c(tmp3, tmp2)
georgeset$biopsy[georgeset$biopsy == "V"] <- NA

georgeset$sample <- 1:ncol(georgeset)
georgeset$platform <- "HG-U133_Plus_2"

#write_rds(georgeset, "../output/georgeset.rds")
```

```{r, echo = FALSE}
```

### So we've gone from this...

```{r}
sapply(eset_lst, function(x) colnames(x)[1:5]) %>% data.frame() %>% kable()
```

### To this...

```{r}
pData(georgeset)[160:180,] %>% kable
```

## Calculate most variable 500 probes and plot as MDS

```{r}
mv500 <- mostVar(exprs(georgeset), 500) %>% row.names()
gset_arg <- mdsArrange(exprs(georgeset[mv500, ]))
mds_input <- merge(gset_arg, pData(georgeset), by.x = 'ids', by.y = 0)
```

### FF vs. FFPE?

```{r}
ggplot(mds_input, aes(x = x, y = y, colour = is.freshfrozen)) + geom_point() + theme_pander()
#ggsave("../figs/tmp.png", dpi = 50)
```

### Batch

```{r}
ggplot(mds_input, aes(x = x, y = y, colour = batch)) + geom_point() + theme_pander()
```

### Controls

```{r}
ggplot(mds_input, aes(x = x, y = y, colour = is.control)) + geom_point() + theme_pander()
```

### Biopsy No.

```{r}
ggplot(mds_input, aes(x = x, y = y, colour = biopsy)) + geom_point() + theme_pander()
```

### Patient Pairs

```{r}
set.seed(123)
ggplot(mds_input[which(mds_input$patient %in% sample(unique(mds_input$patient), 10)), ], 
       aes(x = x, y = y, colour = patient)) + geom_point() + theme_pander()
```

### Technical Replicates

```{r}
ggplot(mds_input[mds_input$is.replicate == TRUE,], 
       aes(x = x, y = y, colour = patient)) + geom_point() + theme_pander()
```

## Are there technical replicates across FF and FFPE samples?

```{r}
xbatch_ids <- georgeset[, georgeset$batch == "A"]$patient[which(georgeset[, georgeset$batch == "A"]$patient %in% georgeset[, georgeset$batch == "B"]$patient)]

ggplot(mds_input[mds_input$patient %in% xbatch_ids, ], aes(x = x, y = y, colour = patient)) + geom_point()
```






