---
title: Unsupervised Analysis
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center', 
                      fig.width = 9)
```

```{r}
library(tidyverse)
library(Biobase)
library(ggthemes)
library(cowplot); theme_set(theme_grey())
source("/Volumes/igmm/sims-lab/Dominic/functions/idReplace.R")
source("/Volumes/igmm/sims-lab/Dominic/functions/mostVar.R")
source("/Volumes/igmm/sims-lab/Dominic/functions/library/mdsArrange.R")
```

#### Here we'll compare the unsupervised approach used with the Edinburgh data with the new Georgetown data. To begin for each set we'll calculate the most variable 500 genes between on-treatment and long-term samples (thereby avoiding the larest likely source of variation - treatment) and test how well they stratify dormany and non-dormant patients.

#### Simlarly as in *subtyping* I'll need to alter ed- or georgeset to match - eventually this will be done properly and comprehensively.

```{r}
edset <- read_rds("../../edinburgh/output/dorm-v4.rds")
edset$xpr_id <- edset$ID_D120days_3cat
edset$patient_id <- edset$patient.no
edset$is_dormant <- edset$dorm.group_v4 == "D"
edset$timepoint <- ifelse(edset$time.point_3cat == 1, "diagnosis", 
                          ifelse(edset$time.point_3cat == 2, "on-treatment", "long-term"))
edset$days_treated <- edset$days_newinfo

georgeset_base <- read_rds("../output/final-georgeset-sep-frma-fselect-loess-clin-cb.Rds")
georgeset_affy <- georgeset_base[, !is.na(georgeset_base$is_dormant)]
xpr_hgnc <- idReplace(exprs(georgeset_affy), 
                        id.in = "affy_hg_u133_plus_2", 
                        id.out = "hgnc_symbol")
phenoData <- new("AnnotatedDataFrame", 
                 data = pData(georgeset_affy),  
                 varMetadata = varMetadata(georgeset_affy))
georgeset <- new("ExpressionSet", 
                   exprs = as.matrix(xpr_hgnc), 
                   phenoData = phenoData)

eset_lst <- list(edset, georgeset)
```

#### We can calculate the 500 most-variable genes independently for each dataset and compare their overlap

```{r}
eset500_lst <- lapply(eset_lst, function(eset){
                          eset_sub <- eset[, which(eset$timepoint == 'long-term')]
                          mv500 <- mostVar(exprs(eset_sub), 500) %>% row.names()
                          eset[mv500,]
})

do.call(intersect, lapply(eset500_lst, row.names))
```

#### as well as how well the are able to cluster patients based on dormancy status

```{r}
plotMDS <- function(eset, time_point = "all", title = ""){
    if(time_point[[1]] != "all"){
               eset_in = eset[, which(eset$timepoint %in% time_point)]
           } else {
               eset_in = eset 
           }
           arng_dfr <- mdsArrange(exprs(eset_in))
           mrg_dfr <- merge(arng_dfr, pData(eset_in), by.x = 'ids', by.y = 'xpr_id')

           p_mds <- ggplot(mrg_dfr, aes(x = x, y = y, colour = is_dormant, size = days_treated)) + 
               geom_point() + 
               labs(title = title,
                    subtitle = paste0("x separation p = ", 
                                      wilcox.test(x~is_dormant, data = mrg_dfr)$p.value)) +
               theme_pander() +
               theme(legend.position = 'top')

           p_box <- ggplot(mrg_dfr, aes(x = is_dormant, y = x, fill = is_dormant)) + 
               geom_boxplot(notch = TRUE) + 
               coord_flip() + 
               theme_pander() + 
               theme(legend.position = 'none')
           plot_grid(p_mds, p_box, rel_heights = c(3, 1), ncol = 1)
}
```

### All samples

```{r, echo = FALSE, eval = FALSE}
#scree plotting and PCA vs MDS comparison
#d <- exprs(eset500_lst[[1]])
#dst <- dist(t(scale(d)))
#dN <- dimnames(d)[[2]]
#dst.m <- as.matrix(dst)
#dimnames(dst.m) <- list(dN, dN)
#dst.cmd <- cmdscale(dst.m, k = 2, eig = TRUE)
#
#eig_dfr <- data.frame(eig = dst.cmd$eig / sum(dst.cmd$eig) * 100, pc = 1:length(dst.cmd$eig))
#
ggplot(dst.cmd, aes(x = pc, y = eig)) + geom_point()

library(ggfortify)
plot(prcomp(t(d)))
dst.cmd <- cmdscale(dst.m, k = 2) %>% as.matrix() %>% data.frame()
ggplot(dst.cmd, aes(X1, X2)) + geom_point()

```{r}
lapply(eset500_lst, function(x) plotMDS(x, 'all', 'All Samples')) %>% 
    plot_grid(plotlist = ., labels = c("Edinburgh", "Georgetown"), scale = 0.9)
```

### Treated samples

```{r}
lapply(eset500_lst, function(x) plotMDS(x, c('on-treatment', 'long-term'), 'Treated Samples')) %>% 
    plot_grid(plotlist = ., labels = c("Edinburgh", "Georgetown"), scale = 0.9)
```

### Long-term samples

```{r}
lapply(eset500_lst, function(x) plotMDS(x, 'long-term', 'Long-term Samples')) %>% 
    plot_grid(plotlist = ., labels = c("Edinburgh", "Georgetown"), scale = 0.9)
```

### On-treatment samples

```{r}
lapply(eset500_lst, function(x) plotMDS(x, 'on-treatment', 'On-treatment Samples')) %>% 
    plot_grid(plotlist = ., labels = c("Edinburgh", "Georgetown"), scale = 0.9)
```

### Diagnostic samples

```{r}
lapply(eset500_lst, function(x) plotMDS(x, 'diagnosis', 'Diagnostic Samples')) %>% 
    plot_grid(plotlist = ., labels = c("Edinburgh", "Georgetown"), scale = 0.9)
```
