---
title: Hallmarks heatmaps
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')
```

<!--
####Setup
-->

```{r}
library(tidyverse)
library(Biobase)
library(GSEABase)
library(cowplot); theme_set(theme_grey())
source("../../lib/arrangeGSEAGenesets.R")
source("../../../../functions/library/heatmapArrange.R")
```

```{r}
dormset <- read_rds("../output/dormset.rds")
hallmarks <- getGmt("../data/h.all.v6.1.symbols.gmt") %>% arrangeGenesets(dormset)

#Significant hallmarks for various comparisons
hm_all <- readLines("../output/hm-gsea-unpaired-all.txt")
hm_diagnosis <- readLines("../output/hm-gsea-unpaired-diagnosis.txt")
hm_on <- readLines("../output/hm-gsea-unpaired-on-treatment.txt")
hm_long <- readLines("../output/hm-gsea-unpaired-long-term.txt")
```

#### Here we're looking to take a geneset that we've determined as informative in _hallmarks-gsea.Rmd_ and display the per-sample expression as a heatmap.

```{r}
heatmapPlot <- function(eset, geneset, title){
        #arrange
        esub <- eset[geneset,]
        dfr <- heatmapArrange(exprs(esub), cluster_row = TRUE, cluster_column = FALSE)
        mrg_dfr <- merge(dfr, pData(esub), by.x = 'variable', by.y = 'xpr_id')
        ord_levels <- unique(mrg_dfr$variable[order(mrg_dfr$days_treated)])
        mrg_dfr$ordered_samples <- factor(mrg_dfr$variable, levels = ord_levels)
        fill_max <- range(mrg_dfr$value) %>% abs() %>% max()

        #plot
        p_hmap <- ggplot(mrg_dfr, aes(x = ordered_samples, y = row_value, fill = value)) +
                        geom_tile() + 
                        #scale_fill_gradient2(high = '#d73027', mid = 'black', low = '#1a9850') +
                        scale_fill_gradientn(values=c(1, .7, .5, .3, 0), 
                                             colours=c("#fb6a4a", "#cb181d", "black", 
                                                       "#238b45", "#74c476"), 
                                             limits = c(-fill_max, fill_max),
                                             na.value = "white") +
                        facet_grid(~is_dormant, scales = 'free_x', space = 'free') + 
                        labs(title = title) +
                        theme(
                              axis.text.x = element_blank(),
                              axis.ticks.x = element_blank()
                              )

        p_bar <- ggplot(mrg_dfr, aes(x = ordered_samples, y = days_treated, colour = timepoint)) +
                        geom_point() + 
                        theme(
                              axis.text.x = element_blank(),
                              axis.ticks.x = element_blank()
                              ) + 
                        facet_grid(~is_dormant, scales = 'free_x', space = 'free')
                        labs(title = title)
        plot_grid(p_bar, p_hmap, ncol = 1, align = 'v', rel_heights = c(1, 5))
}
```

#### All sample comparisons

```{r}
lapply(hm_all, function(x) heatmapPlot(dormset, hallmarks[[x]], x))
```

#### Diagnostic comparisons

```{r}
lapply(hm_diagnosis, function(x) {
               print(x)
               eset <- dormset[, which(dormset$timepoint == "diagnosis")]
               heatmapPlot(eset, hallmarks[[x]], x)
      })
```

#### On-treatment comparisons

```{r}
lapply(hm_on, function(x) {
               eset <- dormset[, which(dormset$timepoint == "on-treatment")]
               heatmapPlot(eset, hallmarks[[x]], x)
      })
```

#### Long-term comparisons

```{r}
lapply(hm_long, function(x) {
               eset <- dormset[, which(dormset$timepoint == "long-term")]
               heatmapPlot(eset, hallmarks[[x]], x)
      })
```


#### From those many, *many* heatmaps, signatures that appear interesting are:
- All samples
  - *Maybe* G2M checkpoint differential expression in long-term samples
  - Estrogen response early dampened in dssns
  - MTORC long-term samples appear differentially regulated
  - E2F targets
  - EMT long-term samples have a region of *more* up-regulated expression as well as a loss of expression 
    unshared in desensitiseds
  - OXPHOS **early** samples express these genes less in desensitiseds - very interesting...
  - Especially when combined with a seeming lack of loss in Glycolysis genes at long-term in dssn
- Diagnostic samples
  - Potential subtle differences in EMT here?
  - Again, OXPHOS appears decreased in dssns
- On-treatment samples
  - Pretty much nothing but this is kind of to be expected
- Long-term samples
  - Potential difference in G2M checkpoint (which matches what was seen in the All Samples plot)
  - Similarly potential differences in Adipogensis
  - Again potential differences in MTORC as noted above
  - And Glycolysis
