---
title: 
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center')
```

```{r}
library(tidyverse)
library(Biobase)
library(ggforce)
```

```{r}
dormset <- read_rds("../output/dormset.rds")
```

#### Here we're going to characterise our patients based on the sampling timepoints - *diagnostic*, *on-treatment* and *long-term* - where we would anticipate desensitised patients to exhibit decreasing correlation at later timepoints, and vice versa for dormant patients.

#### We can also extend this categorical comparison to simply compare correlation vs. time on treatment.

```{r}
corArrange <- function(eset){
    cor_mtx <- cor(exprs(eset))
    diag(cor_mtx) <- NA
    cor_vec <- data.frame(cor = colMeans(cor_mtx, na.rm = TRUE))
    cor_mrg <- merge(cor_vec, pData(eset), by = 0)
    cor_mrg$timepoint <- factor(cor_mrg$timepoint, levels = c("diagnosis", "on-treatment", "long-term"))
    cor_mrg
}

corByStatus <- function(eset){
    lapply(c(TRUE, FALSE), function(logical){
               corArrange(dormset[, which(dormset$is_dormant == logical)])
}) %>% do.call(rbind, .)
}

all_cor <- corByStatus(dormset)

ggplot(all_cor, aes(x = is_dormant, y = cor)) + 
    geom_boxplot(outlier.size = 0, notch = TRUE) + 
    geom_sina()

ggplot(all_cor, aes(x = days_treated, y = cor, colour = is_dormant)) + 
    geom_point() + 
    geom_smooth()


timepoint_cor <- lapply(c("diagnosis", "on-treatment", "long-term"), function(timepoint){
                            corByStatus(dormset[, which(dormset$timepoint == timepoint)])
}) %>% do.call(rbind, .)

ggplot(timepoint_cor, aes(x = timepoint, y = cor)) + 
    geom_boxplot(outlier.size = 0, notch = TRUE) +
#    geom_jitter(width = 0.2) +
    geom_sina() +
    facet_wrap(~is_dormant, nrow = 1)

ggplot(timepoint_cor, aes(x = days_treated, y = cor, colour = is_dormant)) +
    geom_point() +
    geom_smooth()

```

