---
title: Batch effects and correction
author: Dominic Pearce
output:
    github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = 'center', fig.width = 8, fig.height = 6)
```

```{r}
library(tidyverse)
library(Biobase)
library(sva)
library(reshape2)
library(testthat)
library(ggthemes)
library(cowplot); theme_set(theme_gray())
source("../../../../../functions/mostVar.R")
source("../../../../../functions/library/mdsArrange.R")
```

#### We have our frma and loess normalised expression set, but there still clearly exists a big batch difference between FF and FFPE source material. This is evident in terms of the overall sample distributions

```{r}
georgeset <- read_rds("~/Desktop/temp-store/georgeset-sep-frma-fselect-loess-clin.Rds")
```

## Visual determination

### Pre-correction

```{r}
batchEffectDists <- function(mtx_input){
    #arrange
    george_mlt <- melt(mtx_input)
    george_mrg <- base::merge(george_mlt, pData(georgeset), by.x = "Var2", by.y = 0)
    george_mrg$Var2 <- factor(george_mrg$Var2, levels = unique(george_mrg$Var2[order(george_mrg$is_freshfrozen)]))
    #boxplot
    p_box <- ggplot(george_mrg, aes(x = Var2, y = value, fill = is_freshfrozen)) + 
        geom_boxplot() +
        theme(legend.position = 'bottom',
            axis.ticks.x = element_blank(),
            axis.text.x = element_blank())
    #densityplot
    p_dist <- ggplot(george_mrg, aes(x = value, colour = is_freshfrozen, group = Var2), alpha = 0.05) + 
        geom_line(stat = 'density', alpha = 0.2) + 
        theme_pander() +    
        theme(legend.position = 'bottom',
            axis.ticks.x = element_blank(),
            axis.text.x = element_blank()) 
        #combine
    plot_grid(p_box, p_dist, ncol = 1)
}

batchEffectDists(exprs(georgeset))
```

#### ...as well as when comparing just the 500 most variable genes as an mds

```{r}
batchEffectMDS <- function(mtx_input, colour_by){
    #arrange
    mv500 <- mostVar(mtx_input, 500) %>% row.names()
    arg500 <- mdsArrange(mtx_input[mv500,]) 
    mds_input <- base::merge(arg500, pData(georgeset), by.x = 'ids', by.y = 0)
    #and plot
    ggplot(mds_input, aes_string("x", "y", colour = colour_by)) + 
        geom_point() + 
        theme_pander() + 
        theme(legend.position = 'bottom')
}

batchEffectMDS(exprs(georgeset), "is_freshfrozen")
```

#### We can additionally check a number of other variables to check for any other effects

```{r}
var_vec <- c("biopsy", "is_uss", "Age", "ER", "Diagnostic.CB.Grade", "No.of.Pos.Nodes")
plot_lst <- lapply(var_vec, function(variable) batchEffectMDS(exprs(georgeset), variable))
cowplot::plot_grid(plotlist = plot_lst, ncol = 3)
```

#### And as it's perhaps the most likely source of biological variation, we'll plot biopsy number a little larger

```{r}
batchEffectMDS(exprs(georgeset), "timepoint")
```

### Correction and post-correction

#### We'll attempt to correct for this source material-specific batch effect using ComBat

```{r}
g_cb <- ComBat(exprs(georgeset), batch = as.numeric(georgeset$is_freshfrozen))
```

#### And XPN

*note, I've ignored XPN here as it was a bit hacky to get it working and I don't fully trust it*

```{r, eval = FALSE}
#lapply(list.files("~/Desktop/temp-store/inSilicoMerging/R/"), function(file){
#           print(file)
#           source(file.path("~/Desktop/temp-store/inSilicoMerging/R", file))
#            })
#source("~/Desktop/temp-store/inSilicoMerging/R/xpn.R")
#source("~/Desktop/temp-store/inSilicoMerging/R/util.R")
#
##g_xpn <- merge(list(georgeset[, georgeset$is_freshfrozen], georgeset[, !georgeset$is_freshfrozen]), method = "XPN")
#aux <- xpn(georgeset[, georgeset$is_freshfrozen], georgeset[, !georgeset$is_freshfrozen])
#g_xpn <- a4Base::combineTwoExpressionSet(aux[[1]], aux[[2]])
##write_rds(aux, "~/Desktop/temp-store/gset-frma-feature-loess-xpn.rds")
```

ComBat

```{r}
batchEffectDists(g_cb)
```

XPN

```{r}
#batchEffectDists(exprs(g_xpn))
```

#### Better? What about by MDS?

ComBat

```{r}
batchEffectMDS(g_cb, "is_freshfrozen")
batchEffectMDS(g_cb, "timepoint") 
```

XPN

```{r}
#batchEffectMDS(exprs(g_xpn), "is_freshfrozen")
```

#### Prior to feature selecting, the FF vs FFPE effect was still very evident post-correction but now the clear divide caused by source material has seemingly all but been removed.

#### Let's now check the FF-FFPE replicate samples only

```{r}
material_rep_ids <- c("50-1", "50-2", "50-3", "50-4", "135-1", "135-4", "188-1", "268-1", "268-2", "298-1", "298-2", "347-3", "413-1", "413-2", "416-1", "416-2")
material_rep_vec <- rep(FALSE, ncol(georgeset))
material_rep_vec[which(georgeset$sample_id %in% material_rep_ids)] <- TRUE
georgeset$is_material_rep <- material_rep_vec

batchEffectDists(g_cb[, material_rep_vec])
batchEffectMDS(g_cb[, material_rep_vec], "sample_id") + geom_point(aes(size = is_freshfrozen))

#batchEffectDists(g_xpn[, material_rep_vec])
#batchEffectMDS(g_xpn[, material_rep_vec], "is_freshfrozen")
```

#### And write out final .Rds to be used in the actual analysis

```{r}
georgeset_cb <- georgeset
exprs(georgeset_cb) <- g_cb

#test we haven't broken our expressionset as can happen!
test_that("eset overwrite worked as expected", {
              expect_identical(g_cb, exprs(georgeset_cb))
             expect_identical(pData(georgeset), pData(georgeset_cb))
           })

#write_rds(georgeset_cb, "../../output/final-georgeset-sep-frma-fselect-loess-clin-cb.Rds")
```

## Statistical determination 

We can losely define the batch-sepcific *inter*-patienty and *intra*-patient heterogeneity as biological variation, just as the replicate sample variation can be almost entirely attributed to technical variation or noise. Additionally, the overall variability will be explained by a combination of the above.

Therefore, post-ComBat, if the 'biological' variation remains constant post-ComBat, whislt the technical variation is removed we can be confident that our correction has been effective but not to the degree of removing bioloical signal.

```{r}
getSampleCors <- function(mtx){
    cor_mtx <- cor(mtx)
    diag(cor_mtx) <- NA
    colMeans(cor_mtx, na.rm = TRUE)
}

input_lst <- list(raw_all = exprs(georgeset), 
                  raw_ff = exprs(georgeset)[, georgeset$is_freshfrozen],
                  raw_ffpe = exprs(georgeset)[, !georgeset$is_freshfrozen],
                  raw_rep = exprs(georgeset)[, georgeset$is_material_rep],
                  cb_all = exprs(georgeset_cb), 
                  cb_ff = exprs(georgeset_cb)[, georgeset_cb$is_freshfrozen],
                  cb_ffpe = exprs(georgeset_cb)[, !georgeset_cb$is_freshfrozen],
                  cb_rep = exprs(georgeset_cb)[, georgeset_cb$is_material_rep]
                  )

cor_dfr <- lapply(names(input_lst), function(x){
                      data.frame(cor = getSampleCors(input_lst[[x]]), comparison = x)
                  }) %>% do.call(rbind, .)
cor_dfr$correction <- factor(ifelse(grepl("raw", cor_dfr$comparison), "RAW", "CB"), 
                             levels = c("RAW", "CB"))

ggplot(cor_dfr, aes(x = comparison, cor)) + 
    geom_boxplot(outlier.size = 0) + 
    geom_jitter(width = 0.1, alpha = 0.7) +
    facet_wrap(~correction, nrow = 1, scales = 'free_x') + 
    labs(x = "Comparison Type", 
         y = "Correlation", 
         title = "Pre- and post-correction correlation") +
    theme_pander() +
    theme(panel.spacing = unit(5, "lines"))
```

Happily we can see that whilst our overall and replicate correlations have drastically improved, this has not been at the expense of our batch-specific correlations - i.e. biological variation has been preserved.

```{r}
lapply(names(input_lst), function(x){
           data.frame(Correlation = mean(cor_dfr[cor_dfr$comparison == x,]$cor), 
                      Comparison = x) 
         }) %>% 
    do.call(rbind, .) %>%
    t() %>% 
    knitr::kable()
```



